Data Analysis Final Project

Descriptive Statistics
Continuous variables
```{r}
NHANES2017_18a <- read.csv("C:/Users/paypa/Documents/clean_data")
```

```{r}
library(dplyr)
library(tidyr)
NHANES2017_18a%>%
  drop_na(Moderate)%>%
  filter(Moderate<541)%>%
  summarize(mean(Moderate), sd(Moderate), count=n())
NHANES2017_18a%>%
  drop_na(DR1TTFAT)%>%
  summarize(mean(DR1TTFAT), sd(DR1TTFAT), count=n())
NHANES2017_18a%>%
  drop_na(LDLCholesterol)%>%
  summarize(mean(LDLCholesterol), sd(LDLCholesterol), count=n())
```
Descriptive Statistics for ordinal and nominal variables
```{r}
library(questionr)
freq(NHANES2017_18a$BMIcat, cum=TRUE)
freq(NHANES2017_18a$Agecat, cum=TRUE)
freq(NHANES2017_18a$Gender, cum=TRUE)
freq(NHANES2017_18a$Race, cum=TRUE)
freq(NHANES2017_18a$Diabetes, cum=TRUE)
```

Now generate graphs that represent these descriptive statistics (Continuous)
```{r}
library(ggplot2)
ggplot(NHANES2017_18a, aes(x=LDLCholesterol)) + geom_histogram(bins = 30, color = "white", fill = "lightpink") + labs(title = "LDL Cholesterol Frequency", x="LDL-C", y="count")
ggplot(NHANES2017_18a, aes(x=DR1TTFAT)) + geom_histogram(bins = 30, color = "white", fill = "lightpink") + labs(title = "Daily Fat Intake Frequency", x="Daily fat intake in grams", y="count")

```


Created a new database to filter Moderate activity and change the range
now called NHANES_Moderate_reduced
```{r}
NHANEs_Moderate_reduced <- NHANES2017_18a%>%
  filter(Moderate<541)
  ggplot(NHANEs_Moderate_reduced, aes(x=Moderate)) + geom_histogram(bins = 30, color = "white", fill = "lightpink") + labs(title = "Daily Moderate Physcial Activity Frequency", x="Daily Moderate Physical activity in minutes", y="count")
```

Create a graph for descriptive stats of nominal and ordinal variables
BMI categories
```{r}
BMIcat_table <- with(NHANES2017_18a, table(BMIcat))
prop.table(BMIcat_table)
BMIcat_barchart <- as.data.frame(prop.table(BMIcat_table)*100)
BMIcat_barchart
ggplot(BMIcat_barchart, aes(BMIcat,Freq)) +geom_col(fill="lightpink")
```

Age Categories
```{r}
Agecat_table <- with(NHANES2017_18a, table(Agecat))
prop.table(Agecat_table)
Agecat_barchart <- as.data.frame(prop.table(Agecat_table)*100)
Agecat_barchart
ggplot(Agecat_barchart, aes(Agecat,Freq)) +geom_col(fill="lightpink")
```

Gender
```{r}
Gender_table <- with(NHANES2017_18a, table(Gender))
prop.table(Gender_table)
Gender_barchart <- as.data.frame(prop.table(Gender_table)*100)
Gender_barchart
ggplot(Gender_barchart, aes(Gender,Freq)) +geom_col(fill="lightpink")
```

Race
```{r}
Race_table <- with(NHANES2017_18a, table(Race))
prop.table(Race_table)
Race_barchart <- as.data.frame(prop.table(Race_table)*100)
Race_barchart
ggplot(Race_barchart, aes(Race,Freq)) +geom_col(fill="lightpink")
```

Diabetes
```{r}
Diabetes_table <- with(NHANES2017_18a, table(Diabetes))
prop.table(Diabetes_table)
Diabetes_barchart <- as.data.frame(prop.table(Diabetes_table)*100)
Diabetes_barchart
ggplot(Diabetes_barchart, aes(Diabetes,Freq)) +geom_col(fill="lightpink")
```

Run a T-test
Comparing Cholesterol levels across gender
```{r}
t.test(LDLCholesterol~Gender, var.equal=TRUE, data=NHANES2017_18a)
```
Null Hypothesis: There is no difference between LDL cholesterol for males and females.
Alternative Hypothesis: There is a difference between males and females LDL cholestereol levels.
This t-test shows that the difference between LDL Cholesterol levels for males and females is not statistically significant.

Chi-Square test
```{r}
unique(sort(NHANES2017_18a$LDLCholesterolcat))
LDLC_Race_table <- xtabs(~LDLCholesterolcat + Race, data=NHANES2017_18a)
print(LDLC_Race_table)
prop.table(LDLC_Race_table)*100
chisq.test(LDLC_Race_table)
```
x-squared= 4.85 
df=5 
p-value= 0.44 not significant
There are not any statistically significant differences in this comparison.


Kruskal-Wallis Test with Dunn test
Comparing Age cat and LDL cholestereol leve;s
```{r}
library(dunn.test)
NHANES2017_18a%>%
  drop_na(LDLCholesterol)%>%
  group_by(Agecat)%>%
  summarize(mean(LDLCholesterol), sd(LDLCholesterol), count=n())
#2. Run the Kruskal-Wallis and Dunn tests:
dunn.test(x=NHANES2017_18a$LDLCholesterol, g=NHANES2017_18a$Agecat, kw=TRUE)
```
Alternative Hypothesis: There is a difference between Age categories when comparing across LDL-C levels
Null Hypothesis: There is no difference when comparing Age categories across LDL-D levels
The Kruskal-Wallis test shows us that there is a statistically significant difference across age categories when comparing LDL-C levels because the p-value is 0.
Every single comparison is statistically significant, except for 50-79 year-olds in comparison with 20-39 year-olds.



Correlation and Scatterplot
Comparing Total fat intake to LDL-C
```{r}
library(dplyr)
library(ggplot2)

#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=LDLCholesterol, y=DR1TTFAT), data=NHANES2017_18a) + geom_point(color="lightpink") + geom_smooth(method=lm, color="blue") + labs(x="LDL Cholesterol Levels", y="Daily Fat Intake in grams") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(NHANES2017_18a$LDLCholesterol, NHANES2017_18a$DR1TTFAT, method = "pearson")
```
Alternative hypothesis: There is positive correlation between daily fat intake and LDL-C levels
Null hypothesis: There is no correlation between daily fat intake and LDL-C levels.
Correlation coefficient= -0.038
p-value= 0.05335
The p-value tells us that this is not statistically significant.
There is a slightly negative correlation between daily fat intake and LDL-C levels.

Linear Regression
LDL-C by moderate
```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)
Moderate_LDLC_Fit <- lm(LDLCholesterol~Moderate, data=NHANES2017_18a)
summary(Moderate_LDLC_Fit)
```
What is the regression equation for this model?
y= 109.58 + (-0.009)x
What is the multiple R-squared value for this equation?
0.0002
What is the p-value?
0.66
What do these results indicate about the relationship between LDL-C and Moderate Daily Exercise?
The multiple R-squared value tells us that there is likely no correlation between LDL-C levels and Moderate Daily Exercise. This is also not significant because the p-value is 0.66.


Odds Ratio
```{r}
library(epitools)
library(ggmosaic)
#1. Create the odds table:
#odds_table_name <- table(data$IndVar, data$DepVar)
LDLC_Race_Odds_Table <- table(NHANES2017_18a$Race, NHANES2017_18a$LDLCholesterolcat)
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_Race_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(Race), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by Race", x="Race", y="LDL-C levels")
```

Questions: Compared to Mexican Americans, what are the odds ratios having optimal LDL-C levels for each racial category?
Compared to Mexican Americans, multi-race individuals are 0.65 times more likely to have optimal LDL-C. Compared to Mexican Americans, non-Hispanic Asian individuals are 0.77 times more likely to have optimal LDL-C. Compared to Mexican Americans, non-Hispanic Black individuals are 1.03 times more likely to have optimal LDL-C. Compared to Mexican Americans, non-Hispanic white individuals are 0.95 times more likely to have optimal LDL-C. Compared to Mexican Americans, individuals who identify as other are 0.73 times more likely to have optimal LDL-C.
Are these odds ratios significant?
Multi-race p-value= 0.18 not significant
non-Hispanic Asian p-value= 0.28 not significant
non-Hispanic Black p-value= 0.90 not significant
non-Hispanic white p-value= 0.81 not significant
Other p-value= 0.29 not significant


```{r}
LDLC_Gender_Odds_Table <- table(NHANES2017_18a$Gender, NHANES2017_18a$LDLCholesterolcat)
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_Gender_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(Gender), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by Gender", x="Gender", y="LDL-C levels")
```

Questions: Compared to Males, what are the odds ratios having optimal LDL-C levels for each gender category?
Compared to males, females are 0.98 times more likely to have optimal LDL-C.
Are these odds ratios significant?
Female p-value= 0.93 not significant

```{r}
LDLC_Diabetes_Odds_Table <- table(NHANES2017_18a$Diabetes, NHANES2017_18a$LDLCholesterolcat)
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_Diabetes_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(Diabetes), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by Diabetes", x="Diabetes", y="LDL-C levels")
```

Questions: Compared to people who have been told they have borderline diabetes, what are the odds ratios having optimal LDL-C levels for each diabetes category?
Compared to borderline, individuals who have never been told they have diabetes are 1.49 times more likely to have optimal LDL-C.
Compared to borderline, individuals who have been told they have diabetes are 3.79 times more likely to have optimal LDL-C.
Are these odds ratios significant?
No p-value= 0.26 not significant
Yes p-value= 0.0005 significant

Odds Ratio for Moderate categories with LDL-C
```{r}
LDLC_Moderate_Odds_Table <- table(NHANES2017_18a$Moderatecat, NHANES2017_18a$LDLCholesterolcat)
LDLC_Moderate_Odds_Table
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_Moderate_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(Moderatecat), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by Moderate Daily Exercise", x="Moderate Daily Exercise", y="LDL-C levels")
```

Questions: Compared to people who do moderate exercise for less than 30 minutes a day, what are the odds ratios having optimal LDL-C levels for each moderate exercise category?
Compared to <30, individuals who do moderate exercise 120-239 minutes daily are 1.20 times more likely to have optimal LDL-C.
Compared to <30, individuals who do moderate exercise 240+ minutes daily are 1.04 times more likely to have optimal LDL-C.
Compared to <30, individuals who do moderate exercise 30-59 minutes daily are 1.05 times more likely to have optimal LDL-C.
Compared to <30, individuals who do moderate exercise 60-119 minutes daily are 0.99 times more likely to have optimal LDL-C.
Are these odds ratios significant?
120-239 p-value= 0.43 not significant
240+ p-value= 0.93 not significant
30-59 p-value= 0.78 not significant
60-119 p-value= 0.97 not significant

Odds Ratio for BMI cat with LDL-C
```{r}
LDLC_BMIcat_Odds_Table <- table(NHANES2017_18a$BMIcat, NHANES2017_18a$LDLCholesterolcat)
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_BMIcat_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(BMIcat), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by BMI Category", x="BMI Category", y="LDL-C levels")
```

Questions: Compared to people who have a healthy BMI, what are the odds ratios having optimal LDL-C levels for each BMI category?
Compared to healthy, obese individuals are 0.68 times more likely to have optimal LDL-C.
Compared to healthy, overweight individuals are 0.54 times more likely to have optimal LDL-C.
Compared to healthy, underweight individuals are 2.45 times more likely to have optimal LDL-C.

Are these odds ratios significant?
obese p-value= 0.02 significant
overweight p-value= 0.0003 significant
underweight p-value= 0.12 not significant


Odds Ration for LDL-C across Age categories
```{r}
LDLC_Agecat_Odds_Table <- table(NHANES2017_18a$Agecat, NHANES2017_18a$LDLCholesterolcat)
#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)
oddsratio(LDLC_Agecat_Odds_Table)
#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")
ggplot(data=NHANES2017_18a) + geom_mosaic(aes(x=product(Agecat), fill=LDLCholesterolcat), na.rm=TRUE, show.legend=FALSE) + labs(title="LDL-C Levels by Age Category", x="Age Category", y="LDL-C levels")
```

Questions: Compared to people who are less than 20 years old , what are the odds ratios having optimal LDL-C levels for each age category?
Compared to <20, individuals between 20-39 years old are 0.60 times more likely to have optimal LDL-C.
Compared to <20, individuals between 40-59 years old are 0.27 times more likely to have optimal LDL-C.
Compared to <20, individuals between 50-79 years old are 0.59 times more likely to have optimal LDL-C.
Compared to <20, individuals 80+ years old are 1.03 times more likely to have optimal LDL-C.
Are these odds ratios significant?
20-39 p-value= 0.25 not significant
40-59 p-value= 0.002 significant
50-79 p-value= 0.22 not significant
80+ p-value= 0.96 not significant